---
title: "Mediation analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyr)
library(readr)
library(dplyr)

library(ggplot2)
library(cowplot)
```

```{r message=F}
# set path for pre-calculated data, outside the code repo
# `local` / `remote` (reading data from RDSF)
currently_working_env = "local"
source("set_paths.R")
set_paths(currently_working_env)

# metafile
data_lookup<-read_csv(paste0("metadata/data_lookup.csv")) 


source("functions.R")

```



### Univariable MR:  BMI -> Breast cancer (total effect)

```{r, message=F}

#total effects file
total_effects <- read_csv("../02_Results/univariableMR_BMI_BCACall.tsv") %>% filter(exposure == 'Childhood BMI')

# specify which outcome we're going to process
outcome_name <- 'BCAC_full' # see outcome names in total_effects
uni_BC <- total_effects %>% filter(outcome == outcome_name)

exposure_total_beta <- uni_BC %>% pull(b)
exposure_total_se <- uni_BC %>% pull(se)
```



# Mediation analysis

Mediation analysis can be used to dissect the total causal effect of the risk factor on the outcome into an indirect effect of the risk factor on the outcome via the mediator, and a direct effect of the risk factor on the outcome not via the mediator.


## Difference method

INDIRECT = TOTAL EFFECT (of exposure, univ) - DIRECT EFFECT (of exposure, mvmr in presence of the mediator)

In difference method we use Propagation of Errors (PoE) method to calculate SE 

```{r}
# function
difference_method_PoE <- function(total_beta, total_se, direct_beta, direct_se, verbose = F){
  # calculate indirect effect of exposure on outcome (via mediator) 
  # i.e. how much mediator accounts for total effect of exposure on outcome effect
  
  # calculate indirect effect beta
  # INDIRECT = TOTAL (of exposure, univ) - DIRECT (of exposure, mvmr)
  indirect_beta = total_beta -  direct_beta
  if (verbose) {
    print(paste("Indirect effect = ", round(total_beta, 2)," - ", round(direct_beta, 2),   " = ", round(indirect_beta,2)))
  }
  
  
  # calculate SE of indirect effect
  ### using propagation of errors method
  # SE of INDIRECT effect (difference) = sqrt(SE TOTAL^2 + SE DIRECT^2) 
  indirect_se = round(sqrt(total_se^2 + direct_se^2), 4)
  if (verbose) {
    print(paste("SE of indirect effect = sqrt(",round(total_se, 2),"^2 + ", round(direct_se,2),  "^2) = ", indirect_se))
  }
  
  # put data into a tidy df
  df <-data.frame(b= indirect_beta,
                  se = indirect_se)
  
  # calculate CIs 
  df$lo_ci    <- df$b - 1.96 * df$se
  df$up_ci    <- df$b + 1.96 * df$se

  #df<-round(df,3)
  return(df)
}
```


## Product method 

2 ways of calculating the indirect estimate

INDIRECT = TOTAL EFFECT (exposure -> mediator) x TOTAL EFFECT (mediator -> outcome)     (i.e step 1 x step2 from 2-step MR)
INDIRECT = TOTAL EFFECT (exposure -> mediator) x DIRECT EFFECT (of mediator on outcome, from mvmr) 

2 ways of calculating SE:

1) Delta
2) Propagation of errors

```{r}
# function
product_method_Delta <- function(EM_beta, EM_se, MO_beta, MO_se, verbose=F){
  # calculate indirect effect of exposure on outcome (via mediator) 
  # i.e. how much mediator accounts for total effect of exposure on outcome effect

  # this function can run either of two method, depending on what MO data df you supply 
  
  # method 1
  # INDIRECT = TOTAL (exposure -> mediator) x TOTAL (mediator -> outcome)
  # method 2
  # INDIRECT = TOTAL (exposure -> mediator) x DIRECT (of mediator , mvmr) 
  
  # in both cases, it's just EM x MO betas multiplication
  
  
  # calculate indirect effect beta
  EO_beta <- EM_beta * MO_beta
  
  if (verbose) {
    print(paste("Indirect effect = ", round(EM_beta, 2)," x ", round(MO_beta,2), " = ", round(EO_beta, 3)))
  }
  
  
  # Calculate indirect effect SE using Delta method (aka Sobel test) 
  EO_se = sqrt( (MO_beta^2 * EM_se^2) + (EM_beta^2 * MO_se^2) )
  
  # put data into a tidy df
  df <-data.frame(b = EO_beta,
                  se = EO_se,
                  lo_ci = EO_beta - 1.96 * EO_se,
                  up_ci= EO_beta + 1.96 * EO_se)
  
  if (verbose) {
    print(paste("SE of indirect effect = ", round(df$se, 2)))
  }

  #df<-round(df,3)
  return(df)
}

product_method_PoE <- function(EM_beta, EM_se, MO_beta, MO_se, verbose=F){
  # calculate indirect effect of exposure on outcome (via mediator) 
  # i.e. how much mediator accounts for total effect of exposure on outcome effect

  # this function can run either of two method, depending on what MO data df you supply 
  
  # method 1
  # INDIRECT = TOTAL (exposure -> mediator) x TOTAL (mediator -> outcome)
  # method 2
  # INDIRECT = TOTAL (exposure -> mediator) x DIRECT (of mediator , mvmr) 
  
  # in both cases, it's just EM x MO betas multiplication
  
  # calculate indirect effect beta
  EO_beta <- EM_beta * MO_beta
  
  if (verbose) {
    print(paste("Indirect effect = ", round(EM_beta, 2)," x ", round(MO_beta,2), " = ", round(EO_beta, 3)))
  }
  
  
  # calculate SE of indirect effect 
  ### using propagation of errors method
  # SE of INDIRECT effect (difference) = sqrt(SE EM^2 + SE MO^2) 
  EO_se = round(sqrt(EM_se^2 + MO_se^2), 4)
  if (verbose) {
    print(paste("SE of indirect effect = sqrt(",round(EM_se, 2),"^2 + ", round(MO_se,2),  "^2) = ", EO_se))
  }
  
  # put data into a tidy df
  df <-data.frame(b= EO_beta,
                  se = EO_se)
  
  # calculate CIs 
  df$lo_ci    <- df$b - 1.96 * df$se
  df$up_ci    <- df$b + 1.96 * df$se

  #df<-round(df,3)
  return(df)
}
```



# Load data for all mediators and store all required betas for mediation analysis in a df

```{r, message=F}
# mediators to load
mediators_df <- data_lookup %>%
  filter(trait_category %in% c('sleep_traits')) %>%
  filter(!is.na(source) | !is.na(mrbase.id)) %>% 
  select(trait_file_name, trait, trait_category) 

# create empty storage df
collect_betas <-data.frame(matrix(ncol=9,
                dimnames=list(NULL, c("mediator_name",
                                      "total_exposure_beta","total_exposure_se",
                                      "total_mediator_beta", "total_mediator_se",
                                      "direct_exposure_beta", "direct_exposure_se",
                                      "direct_mediator_beta", "direct_mediator_se"))))

for (mediator in mediators_df$trait_file_name){
  
  # load all required files for each mediator in the loop
  mediator_name <- mediators_df %>% filter(trait_file_name == mediator) %>% pull(trait)
  current_trait_category <- mediators_df %>% filter(trait_file_name == mediator) %>% pull(trait_category)
  
  # results from step 1
  bmi_mediator <- read_tsv(paste0(results_path, current_trait_category, "/MR_BMI-to-",mediator,".tsv")) %>%
                  filter(outcome == mediator_name, 
                         exposure == "Childhood body size",
                         method == 'Inverse variance weighted')
  
  # results from step 2                          
  mediator_bc <- read_tsv(paste0(results_path, current_trait_category, 
                                 "/merged_mr_mediators_", current_trait_category,"-to-", outcome_name,".tsv")) %>% 
                 filter(method %in% c("Inverse variance weighted", "Wald ratio")) %>% 
                 filter(exposure == mediator_name )
  
  # results from mvmr  
  mvmr<- read_tsv(paste0(results_path, current_trait_category, 
                                 "/mvmr/mvmr_earlyBMI-", mediator ,"-to-", outcome_name,"_using_2SMR.tsv"))
    
  
  
  # store all required data in a vector
  mediator_data<- c(mediator_name,
                    bmi_mediator %>% pull(b)%>% as.numeric(),
                    bmi_mediator %>% pull(se)%>% as.numeric(),
                    mediator_bc %>% pull(b)%>% as.numeric(),
                    mediator_bc %>% pull(se)%>% as.numeric(),
                    mvmr %>% filter(exposure == "Childhood body size" )%>% pull(b),
                    mvmr %>% filter(exposure == "Childhood body size" )%>% pull(se),
                    mvmr %>% filter(exposure != "Childhood body size")%>% pull(b),
                    mvmr %>% filter(exposure != "Childhood body size")%>% pull(se) ) 
  
  # add to storage df
  collect_betas<-rbind(collect_betas, mediator_data) %>% drop_na()
  print(paste0("Done ", mediator_name))
}

# make sure values are in numeric format
collect_betas <- collect_betas %>%
  mutate_at(vars(-mediator_name), function(x) as.numeric(as.character(x)))

```


# Run mediation analysis (3 types) on mediators with collected betas. 
```{r, echo=F}

# make empty results storage df
mediation_results <- data.frame(matrix(ncol=17,
                dimnames=list(NULL, 
                              c("mediator_name",
                                
                                "difference_method_beta", "difference_method_se", 
                                "difference_method_lo_ci95", "difference_method_up_ci95", 

                                "product_method_v1_beta", "product_method_v1_se",
                                "product_method_v1_lo_ci95", "product_method_v1_up_ci95",

                                "product_method_v2_beta", "product_method_v2_se",
                                "product_method_v2_lo_ci95", "product_method_v2_up_ci95",

                                "product_method_v3_beta", "product_method_v3_se",
                                "product_method_v3_lo_ci95", "product_method_v3_up_ci95"))))

# make sure we have a place to write
mvmr_dir <- paste0(results_path, "sleep_traits/mediation/")
mvmr_dir_w_backslash <- gsub(" ", "\\ ", mvmr_dir, fixed=T) # create path vector escaping spaces, otherwise sytem call cant process it
if(!dir.exists(mvmr_dir)){ system(paste("mkdir -p", mvmr_dir_w_backslash))}

# loop to run mediation
for (mediator in collect_betas$mediator_name){
  print(paste0("===== processing ", mediator))
  collect_betas_sub<-collect_betas %>%  filter(mediator_name == mediator)
  
  # difference method
  print(~"Difference method + PoE")
  diff_output <- difference_method_PoE(total_beta = exposure_total_beta,
                                       total_se = exposure_total_se, 
                                       direct_beta = collect_betas_sub$direct_exposure_beta, 
                                       direct_se = collect_betas_sub$direct_exposure_se,
                                       verbose = T)
  
  
  # product method, v1 : using step 1 x step 2
  print(~"Product method (step 1 x step 2) + Delta")
  prod_output1 <- product_method_Delta(EM_beta = collect_betas_sub$total_exposure_beta,
                                       EM_se =   collect_betas_sub$total_exposure_se,
                                       MO_beta = collect_betas_sub$total_mediator_beta,
                                       MO_se =   collect_betas_sub$total_mediator_se,
                                       verbose = T)
  
  
  # product method, v2 : using step 1 x mvmr direct effect          #### this method is likely to create the 'best' results
  print(~"Product method (step 1 x mvmr) + Delta")
  prod_output2 <- product_method_Delta(EM_beta = collect_betas_sub$total_exposure_beta,
                                       EM_se =   collect_betas_sub$total_exposure_se,
                                       MO_beta = collect_betas_sub$direct_mediator_beta,
                                       MO_se =   collect_betas_sub$direct_mediator_se,
                                       verbose = T)
  

  # product method, v3, using PoE method for SE, with  step 1 x mvmr direct effect 
  print(~"Product method (step 1 x mvmr) + PoE")
  prod_output3 <- product_method_PoE(EM_beta = collect_betas_sub$total_exposure_beta,
                                     EM_se =   collect_betas_sub$total_exposure_se,
                                     MO_beta = collect_betas_sub$direct_mediator_beta,
                                     MO_se =   collect_betas_sub$direct_mediator_se,
                                     verbose = T)
  



  # store results required data in a vector
  mediation_results_sub <- c(mediator,
                             
                            diff_output$b, diff_output$se,
                            diff_output$lo_ci, diff_output$up_ci,
                            diff_output$or, diff_output$or_lci95, diff_output$or_uci95,
                            
                            prod_output1$b, prod_output1$se,
                            prod_output1$lo_ci,  prod_output1$up_ci, 
                            prod_output1$or, prod_output1$or_lci95, prod_output1$or_uci95,
                            
                            prod_output2$b, prod_output2$se,
                            prod_output2$lo_ci,  prod_output2$up_ci, 
                            prod_output2$or, prod_output2$or_lci95, prod_output2$or_uci95,
                            
                            prod_output3$b, prod_output3$se,
                            prod_output3$lo_ci,  prod_output3$up_ci, 
                            prod_output3$or, prod_output3$or_lci95, prod_output3$or_uci95)
  

  # add to results storage df
  mediation_results<- rbind(mediation_results, mediation_results_sub) %>% drop_na()
  print(paste0("== Finished  ", mediator))
  
}  
# make sure values are in numeric format
mediation_results <- mediation_results %>%
  mutate_at(vars(-mediator_name), function(x) as.numeric(as.character(x)))


mediation_results %>% 
  write_tsv(paste0(results_path, "sleep_traits/mediation/mediation_analysis_results_", outcome_name,".tsv"))
```



# Plot all results

```{r}

# read the file produced before
mediation_results <-  read_tsv(paste0(results_path, "sleep_traits/mediation/mediation_analysis_results_", outcome_name,".tsv"))

## PLOT OF betas WITH CIs
  
mediation_results_long <- mediation_results %>%
  select("mediator_name", !contains(c("or", "se"))) %>%  
  pivot_longer(cols = c( "product_method_v1_beta" , "product_method_v2_beta" ,  "product_method_v3_beta" ,"difference_method_beta"), names_to = "beta", values_to = "indirect_beta") %>%
  pivot_longer(cols = c( "product_method_v1_lo_ci95" , "product_method_v2_lo_ci95" ,  "product_method_v3_lo_ci95" ,"difference_method_lo_ci95"), names_to = "lo_ci95", values_to = "indirect_lo_ci95") %>% 
  pivot_longer(cols = c( "product_method_v1_up_ci95" , "product_method_v2_up_ci95" ,"product_method_v3_up_ci95" , "difference_method_up_ci95"), names_to = "up_ci95", values_to = "indirect_up_ci95") %>% 
  mutate(beta = gsub("_beta", "", beta)) %>% 
  mutate(lo_ci95 = gsub("_lo_ci95", "", lo_ci95)) %>%
  mutate(up_ci95 = gsub("_up_ci95", "", up_ci95)) %>% 
  mutate(method = ifelse(beta == lo_ci95 & lo_ci95 == up_ci95, 1, 0)) %>% 
  filter(method == 1) %>% 
  select(-lo_ci95, -up_ci95, -method) %>% 
  dplyr::rename("mediation_type" = "beta")



  mediation_results_long_out<- mediation_results_long %>% 
    # record SE method
    mutate(standard_error= case_when(mediation_type %in% c('difference_method', 'product_method_v3') ~ 'PoE',
                                     mediation_type %in% c('product_method_v1', 'product_method_v2') ~ 'Delta')) %>% 
    # add nice names to types of mediation analysis
    mutate(mediation_type= case_when(mediation_type == "product_method_v1" ~ "Product method V1 (total)", # can add more specific names
                                  mediation_type == "product_method_v2" ~    "Product method V2 (direct)",
                                  mediation_type == "product_method_v3" ~    "Product method V3 (direct)",
                                  mediation_type == "difference_method" ~    "Difference method", 
                                  TRUE ~ mediation_type)) %>% 
    # set the order in which mediation methods appear in the plot
    mutate(mediation_type= factor(mediation_type, 
                                  levels = rev(c( "Product method V1 (total)", 
                                              "Product method V2 (direct)",
                                              "Product method V3 (direct)",
                                              "Difference method" )))) %>% 
    # set the order in which mediaotrs appear in the plot
    mutate(mediator_name = factor(mediator_name ,
                                  levels = c("Chronotype"))) # add here other mediator names, in the order you want them to appear in the plot

  
# (optional: SKIP THIS at the first run - use this after you decide which methods to keep)
# only display selected methods method
  
#mediation_results_long_out <- mediation_results_long_out %>% 
#  filter(mediation_type %in% c("Product method V2 (direct)", "Difference method"))




# NB this is a heavily modified version for pulication
p<-ggplot(mediation_results_long_out,
           aes(y=mediator_name, x=indirect_beta, 
               label=mediation_type , group = mediation_type, colour=mediation_type, shape=standard_error)) +
  geom_vline(xintercept=0, linetype=2, color= "grey50") +
  geom_point(size=2,  
             position = ggstance::position_dodgev(height = 0.5))+
  geom_errorbarh(aes(xmin=indirect_lo_ci95, xmax=indirect_up_ci95), height=.2, 
            position = ggstance::position_dodgev(height = 0.5)) +
  theme_minimal_grid(font_size =8, line_size = 0.6) +
  #scale_color_manual(values=pal)+  # can provide a colour palette
  scale_shape_manual(values = c(20,1)) +
  theme(strip.text = element_text(face = 'bold'),
        panel.grid.major.y = element_blank())+
  scale_x_continuous(breaks=seq(-0.2,0.2,0.1))+  # NB change this if your CIs are wider

  labs(color = "Mediation method",
       shape = "Standard error method",
       y = "", x = "beta coefficient of indirect effect",
       title= paste0("Mediation analysis: indirect effect of childhood body size via mediators on ", outcome_name)
       )+
  theme(legend.position = "right",  plot.title.position = "plot")+
  guides(color = guide_legend(order=1, reverse=TRUE),
         shape = guide_legend(order=2))
p

# save the plot: may need to adjust heigh/width
ggsave(paste0(results_path, "figures/mediation_analysis_", outcome_name,".png"),
       plot=p, scale=1.2, 
       width=14, height=6,
       units=c("cm"), dpi=300, limitsize=F)

```

```{r, eval=F}
#### IGNORE THIS
# there are several variation of the mediation results plot - 
# some version are below, but they may not work or need to modified to work with this data
# we can discuss this later



b2<-ggplot(mediation_results_long_out,
           aes(y=mediator_name, x=indirect_beta, 
               label=mediation_type , group = mediation_type, colour=mediation_type)) +
  geom_point(size=3, shape=20, 
             position = ggstance::position_dodgev(height = 0.5))+
  geom_errorbarh(aes(xmin=indirect_lo_ci95, xmax=indirect_up_ci95), height=.1, 
            position = ggstance::position_dodgev(height = 0.5)) +
  theme_minimal_grid(9) +
  #scale_color_manual(values=pal)+
  geom_vline(xintercept=0, linetype=3) +
  facet_wrap(~mediation_type, ncol=1)+
  scale_x_continuous(breaks=seq(-0.2,0.4,0.1))+
  theme(strip.text = element_text(face = 'bold'))+
  labs(color = "",y = "", x = "Effect size (beta)", 
       subtitle="                   estimated using two mediation analysis methods: ",
       title= "              Mediation analysis: indirect effect of childhood BMI via mediator" )+
  theme(legend.position = "none",  plot.title.position = "plot")


b1v2<-ggplot(mediation_results_long_out,
           aes(y=mediator_name, x=indirect_beta, 
               label=mediation_type , group = mediation_type, colour=mediation_type)) +
  geom_vline(xintercept=0, linetype=2, color= "grey50") +
  geom_point(size=2, 
             position = ggstance::position_dodgev(height = 0.5))+
  geom_errorbarh(aes(xmin=indirect_lo_ci95, xmax=indirect_up_ci95), height=.2, 
            position = ggstance::position_dodgev(height = 0.5)) +
  theme_minimal_grid(11) +
  #scale_color_manual(values=c("#DA668B","#6A115A"))+
  #scale_color_manual(values=pal)+
  coord_flip()+
  theme(strip.text = element_text(face = 'bold'),
        panel.grid.major.y = element_blank(),
        axis.text.x = element_text(angle = 25, hjust = 1))+
  scale_x_continuous(breaks=seq(-0.2,0.2,0.1))+
  labs(color = "",
       shape = "Standard error method",
       y = "", x = "Effect size (beta)", 
       subtitle="                indirect effect of childhood body size via mediator ",
       title= "            Mediation analysis" )+
  theme(legend.position = "top",  plot.title.position = "plot")


```



